---
title: "qgtools"
output: github_document
---

## Overview

**qgtools** is a framework for specifying and fitting mixed and hierarchical models, with a focus on quantitative genetics and complex covariance structures.

The key idea behind qgtools is to **separate model specification from model fitting**, allowing the same model to be estimated using different software backends and estimation paradigms within a unified modeling framework.


## Core ideas

- **Formulas define model structure**  
  Per-trait formulas specify responses, fixed effects, and random effects, but not how covariance is modeled.

- **Variance components define covariance**  
  Variance components describe how random effects induce covariance across individuals and traits, using kernels and optional priors.

- **Kernels are reusable objects**  
  Pedigree, genomic, and other kernels are standalone objects that can be reused across models and fitting tasks.

- **Tasks control estimation**  
  Models are fitted using `gfit()` with `task = "reml"`, `"bayes"`, or `"solve"`, without changing the model specification.

- **Scalable data access**  
  Data are treated as a data source, allowing both in-memory and disk-backed datasets to be used transparently.

## R and Python interfaces

qgtools is designed with a language-agnostic core, allowing both **R and Python**
interfaces to interact with the same underlying computational backends.

Model specification and orchestration can be performed in either R or Python,
while computationally intensive tasks are handled by shared C++/Fortran libraries.
This design ensures consistent results across interfaces and enables flexible
deployment in cloud and HPC environments.


#### Prepare input data

```{r, eval = FALSE}

## Data may be provided as an in-memory data frame or as a disk-backed file
data <- makeDatalist(
  source = "data.txt",
  format = "CSV"
)

## Prepare pedigree kernel for additive genetic effects
PED <- makePEDlist(fnPED = "pedigree.txt", method = "S-D-NonInbred")

## Prepare genomic relationship kernel for additive genomic effects
GRM <- makeGRMlist(fnGRM = "grm_inverse.txt", format = "BINARY", grm_type = "G-inverse")

# Marker-level genotype container (e.g. PLINK BED/BIM/FAM). This provides information about where genotype data live on disk and how animals map to markers.
Glist <- makeGlist(
  bedfiles = "chr.bed",
  bimfiles = "chr.bim",
  famfiles = "chr.fam"
)
```

#### Linear mixed models: REML and solver-based estimation

```{r, eval = FALSE}


## Single-trait linear mixed model (REML)

## Model formula:
## (1 | id) represents the additive genetic (animal) effect
formulas <- list(
  BW = BW ~ sex + reps + (1 | dam) + (1 | id)
)

## Variance components define covariance structure
vcs <- list(
  dam_env = vc(
    index  = "dam",
    traits = "BW"
  ),

  animal_genetic = vc(
    index  = "id",
    traits = "BW",
    kernel = PED
  ),

  residual = vc(
    index  = "Residual",
    traits = "BW"
  )
)

## Fit the model using REML
fit <- gfit(formulas, data, vcs, task = "reml")

## Multi-trait linear mixed model (REML)

formulas_mt <- list(
  Gl = Gl ~ sex + reps + (1 | dam) + (1 | id),
  BW = BW ~ sex + reps + (1 | dam) + (1 | id)
)

## Shared variance components induce correlation between traits
## For REML, supplied values are used as starting values
vcs_mt <- list(
  dam_env = vc(
    index  = "dam",
    traits = c("Gl", "BW"),
    start  = matrix(
      c(1.0, 0.2,
        0.2, 2.0),
      nrow = 2,
      byrow = TRUE
    )
  ),

  animal_genetic = vc(
    index  = "id",
    traits = c("Gl", "BW"),
    kernel = PED,
    start  = matrix(
      c(4.0, 0.8,
        0.8, 3.0),
      nrow = 2,
      byrow = TRUE
    )
  ),

  residual = vc(
    index  = "Residual",
    traits = c("Gl", "BW"),
    start  = diag(c(5, 5))
  )
)

## REML estimation
fit_mt <- gfit(formulas_mt, data, vcs_mt, task = "reml")

## Solve mixed model equations (no variance updates)
fit_mt <- gfit(formulas_mt, data, vcs_mt, task = "solve")

```

#### Bayesian hierarchical linear mixed models

```{r, eval = FALSE}


## Single-trait Bayesian linear mixed model

formulas <- list(
  BW = BW ~ sex + reps + (1 | dam) + (1 | id)
)

# Specify prior distrubutions for selected factors
priors <- list(
  dam_env = prior(
    index = "dam",
    traits = "BW",
    distribution = iw(df = 4, S = 1),
    start = 1
  ),

  animal_genetic = prior(
    index = "id",
    traits = "BW",
    kernel = PED,
    distribution = iw(df = 4, S = 1),
    start = 1
  ),

  residual = prior(
    index = "Residual",
    traits = "BW",
    distribution = iw(df = 4, S = 1),
    start = 1
  )
)

# Estimate parameters using sampling based methods such Gibbs
fit <- gfit(formulas, data, priors, task = "bayes")


## Multi-trait Bayesian linear mixed model

formulas_mt <- list(
  Gl = Gl ~ sex + reps + (1 | dam) + (1 | id),
  BW = BW ~ sex + reps + (1 | dam) + (1 | id)
)

priors_mt <- list(
  dam_env = prior(
    index  = "dam",
    traits = c("Gl", "BW"),
    distribution = iw(df = 4, S = diag(2)),
    start  = matrix(
      c(1.0, 0.2,
        0.2, 2.0),
      nrow = 2,
      byrow = TRUE
    )
  ),

  animal_genetic = prior(
    index  = "id",
    traits = c("Gl", "BW"),
    kernel = PED,
    distribution = iw(df = 4, S = diag(2)),
    start  = matrix(
      c(4.0, 0.8,
        0.8, 3.0),
      nrow = 2,
      byrow = TRUE
    )
  ),

  residual = prior(
    index  = "Residual",
    traits = c("Gl", "BW"),
    distribution = iw(df = 4, S = diag(2)),
    start  = diag(c(5, 5))
  )
)

fit_mt <- gfit(formulas_mt, data, priors_mt, task = "bayes")
```

#### Bayesian multi-component linear models with marker-level priors

```{r, eval = FALSE}

## Single-trait Bayesian linear regression with marker-level priors

formulas <- list(
  BW = BW ~ sex + reps + (1 | dam)
)

# Optional marker-level metadata
# Used by marker-based priors (e.g. BayesC / BayesR)
# NULL means default behavior
sets       <- NULL
weights    <- NULL
annotation <- NULL

# Prior specification (Bayesian models only)
priors <- list(
  dam_env = prior(
    index = "dam",
    traits = "BW",
    distribution = iw(df = 4, S = 1),
    start = 1
  ),
  marker_genetic = prior(
    index  = "marker",     # marker-level effects
    traits = "BW",
    kernel = Glist,        # genotype / marker kernel
    distribution = bayesC(
      pi         = c(0.95, 0.05),
      gamma      = c(0, 1),
      sets       = sets,
      weights    = weights,
      annotation = annotation
    ),
    start = 1
  ),

  residual = prior(
    index  = "Residual",
    traits = "BW",
    distribution = iw(df = 4, S = 1),
    start = 1
  )
)

```

#### Example R vs Python interface

```{r, eval = FALSE}

# R interface
formulas <- list(
  BW = BW ~ sex + reps + (1 | id)
)

vcs <- list(
  animal = vc(index = "id", traits = "BW", kernel = PED)
)

fit <- gfit(formulas, data, vcs, task = "reml")

# Python interface
model = Model(
    formulas={
        "BW": "BW ~ sex + reps + (1 | id)"
    },
    vcs=[
        vc(index="id", traits=["BW"], kernel=PED)
    ]
)

fit = gfit(model, data, task="reml")


```


## Performance and deployment

qgtools is designed as a lightweight R (or Python) interface to high-performance computing backends.  
Computationally intensive components (e.g. likelihood evaluation, large linear algebra, and sampling) can be implemented in compiled languages such as **C++ or Fortran**, while R (or Python) is used for model specification and orchestration.

This separation provides:
- high computational performance,
- scalability to large datasets,
- and a clear boundary between model definition and numerical implementation.

For deployment, qgtools can be packaged into **containerized environments** (e.g. Docker or Singularity), allowing models to be executed reproducibly on high-performance computing platforms and cloud infrastructures such as **AWS** and **Azure**. Containers bundle the required libraries and runtimes and can be deployed on HPC batch systems or cloud services without exposing source code.

This design enables qgtools to act as a unifying modeling layer while supporting multiple computational backends and deployment scenarios, from local workstations to large-scale cloud and HPC environments.

**qgtools** handles large-scale data by taking advantage of:  
  
* multi-core processing using [openMP](https://www.openmp.org/)  
* multithreaded matrix operations implemented in BLAS libraries (e.g. [OpenBLAS](https://www.openblas.net/), [ATLAS](https://math-atlas.sourceforge.net/) or [MKL](https://en.wikipedia.org/wiki/Math_Kernel_Library))  
* fast and memory-efficient batch processing of genotype data stored in binary files (e.g. [PLINK](https://www.cog-genomics.org/plink2) bedfiles)  

